Windowed least-squares model reduction for dynamical systems

### 2019-10-09 cavity.tex
updated appendix

\subsection{Cavity Flow}
The next numerical example considers hyper-reduced ROMs of a viscous, compressible flow in a two-dimensional cavity. 
The flow is described by the two-dimensional compressible Navier-Stokes equations,
\begin{equation}\label{eq:compressible_ns}
\frac{\partial \mathbf{u}}{\partial t} + \nabla \cdot \big( \mathbf{F}(\mathbf{u} ) - \mathbf{F}_v (\mathbf{u},\nabla \mathbf{u} )      \big) =\bz,
\end{equation}
where $\mathbf{u}: [0,T] \times \Omega \rightarrow \RR{4}$ comprise the density, x and y momentum, and total energy. The terms $\mathbf{F}$ and $ \mathbf{F}_v$ are the inviscid and viscous fluxes, respectively. For a two-dimensional flow, the state vector and inviscid fluxes are
$$
\mathbf{u} = \begin{Bmatrix}
\rho \\ \rho u_1 \\ \rho u_2 \\ \rho E \end{Bmatrix}, \qquad \mathbf{F}_{1} = \begin{Bmatrix} \rho u_1 \\ \rho u_1^2 +      p \\ \rho u_1 u_2 \\ u_1(E + p) \end{Bmatrix}, 
\qquad \mathbf{F}_{2} = \begin{Bmatrix} \rho u_2 \\ \rho u_1 u_2  \\ \rho u_2^2 + p \\ u_2(E + p) \end{Bmatrix}.
$$
The viscous fluxes are given by
$$
\qquad \mathbf{F}_{v_1} = \begin{Bmatrix} 0 \\ \tau_{11} \\ \tau_{12}  \\ u_j \tau_{j1} + c_p \frac{\mu}{\text{Pr}} \frac{\partial T}{\partial x_1}  \end{Bmatrix}, 
\qquad \mathbf{F}_{v_2} = \begin{Bmatrix} 0 \\ \tau_{21} \\ \tau_{22}  \\ u_j \tau_{j2} + c_p \frac{\mu}{\text{Pr}} \frac{\partial T}{\partial x_2}  \end{Bmatrix}.
$$
We assume a Newtonian fluid, which leads to a viscous stress tensor of the form,
\begin{equation*}
\tau_{ij} = 2\mu S_{ij},
\end{equation*}
where,
\begin{equation*}
 S_{ij} = \frac{1}{2} \big( \frac{\partial u_i}{\partial \mathsf{x}_j} + \frac{\partial u_j}{\partial \mathsf{x}_i} \big) - \frac{1}{3} \frac{\partial      u_k}{\partial \mathsf{x}_i} \delta_{ij}.
\end{equation*}
The Navier-Stokes equations are closed with a constitutive relationship for a calorically perfect gas,
$$p = (\gamma - 1)( \rho E - \frac{1}{2} \rho u_1^2 - \frac{1}{2} \rho u_2^2 \big),$$
where $\gamma = 1.4$ is the heat-capacity ratio.

@@

\subsubsection{Description of FOM and Generation of Trial Space}
The full-order model comprises a discontinuous-Galerkin discretization. The discretization 
is obtained by partitioning the domain into $100$ elements in the flow direction and $40$ elements 
in the wall-normal direction. The solution is represented to third order over each element using tensor product polynomials of order $p=2$, 
resulting in $36000$ unknowns for each conserved variable. Spatial discretization via the discontinuous Galerkin method yields a dynamical system 
of the form,
$$\frac{d \state}{dt} = \mass_{\text{DG}} \velocity_{\text{DG}} (\state),$$
where $\mass_{\text{DG}} \in \RR{N \times N}$ is the DG mass matrix and $\velocity_{\text{DG}}: \RR{N} \rightarrow \RR{N}$ is the DG velocity operator containing 
surface and volume integrals. By the definition of the FOM~\eqref{eq:FOM}, the dynamical system velocity is $\velocity = \mass_{\text{DG}} \velocity_{\text{DG}}$. 
Figure~\ref{fig:cav_mesh} shows the computational mesh. The 
DG method leverages the Rusanov flux at the cell interfaces~\cite{rusanov} and uses the first form of Bassi and Rebay~\cite{br1} for the viscous fluxes. Time integration 
is performed via a third-order strong stability preserving Runge-Kutta method~\cite{ssp_rk3} with a time-step size of $\Delta t = 0.001$.

@@

\subsubsection{Description of Reduced-Order Models}
Hyper-reduced ROMs based on the Galerkin, collocated Least--Squares Petrov--Galerkin, and \methodAcronym\ approaches are 
again considered. Details on the implementation of the methods is as follows:
\begin{itemize}
\item \textbf{Galerkin ROM with Collocation}: The Galerkin ROM with collocation is obtained through Galerkin projection of the FOM. Unless 
otherwise noted, the  Galerkin ROM is evolved in time with the Crank--Nicolson time scheme at a time step of $\Delta t =0.1$.

\item \textbf{LSPG ROM with Collocation}: The LSPG ROM with collocation is built on top of the FOM discretization using the Crank--Nicolson scheme for temporal 
discretization. Unless otherwise noted, the time step for the LSPG ROM is $\Delta t  = 0.1$. As in the previous example, 
the Gauss--Newton method is used to solve the non-linear least-squares problem arising at each time instance. The Gauss--Newton 
algorithm is deemed converged when the gradient norm is less than $1e-4$. All Jacobians are again computed via automatic differentiation.

\item \textbf{\methodAcronymROMs\ with Collocation}: \methodAcronymROMs\ are solved via the direct method. The ROMs use the Crank--Nicolson time-discretization with a time-step size of 
$\Delta t = 0.1$. The Gauss--Newton algorithm is deemed converged when the gradient norm is less than $1e-4$. All Jacobians are computed via automatic differentiation. Uniform quadrature 
weights are used for evaluating the integral in~\eqref{eq:obj_gen_slab}.

@@

\subsubsection{Numerical Results}
We first assess the performance of \methodAcronymROMs\ with varying window sizes. To this end, we consider \methodAcronymROMs\ with uniform window 
sizes of $\Delta T^n \equiv \Delta T = 0.2,0.5,1.0,$ and $2.0$, along with the Galerkin and LSPG ROMs. Results are first considered for basis \#1 as described in 
Table~\ref{tab:rom_basis_details}. For all ROMs, we evolve 
the solution for $t \in [0,100]$. This comprises the same time interval used to construct the trial subspace. First, Figure~\ref{fig:cav_results1a} depicts the evolution of the pressure at the bottom wall in the midpoint of the computational domain, while Figure~\ref{fig:cav_results1b} depicts the evolution of the normalized $\elltwo$ state error of the various reduced-order models. Both the hyper-reduced Galerkin and LSPG ROMs blow up/fail to converge within the first several time units. The \methodAcronymROM\ minimizing the residual over a window of size $\Delta T = 0.2$ also fails to converge. The \methodAcronymROMs\ that minimize the residual over window sizes of $\Delta T \ge 0.5$ are seen to all be stable and accurate; the pressure response is well characterized and the normalized state errors are again less than $10\%$. The most notable discrepancy between the \methodAcronymROM\ and FOM solutions is a slight phase difference.

>>>>

\subsection{Cavity flow}
The next numerical example considers hyper-reduced ROMs of a viscous, compressible flow in a two-dimensional cavity. 
The flow is described by the two-dimensional compressible Navier-Stokes equations,
\begin{equation}\label{eq:compressible_ns}
\frac{\partial \mathbf{u}}{\partial t} + \nabla \cdot \big( \mathbf{F}(\mathbf{u} ) - \mathbf{F}_v (\mathbf{u},\nabla \mathbf{u} )      \big) =\bz,
\end{equation}
where $\mathbf{u}: [0,T] \times \Omega \rightarrow \RR{4}$ comprise the density, x and y momentum, and total energy. The terms $\mathbf{F}$ and $ \mathbf{F}_v$ are the inviscid and viscous fluxes, respectively. For a two-dimensional flow, the state vector and inviscid fluxes are
$$
\mathbf{u} = \begin{Bmatrix}
\rho \\ \rho u_1 \\ \rho u_2 \\ \rho E \end{Bmatrix}, \qquad \mathbf{F}_{1} = \begin{Bmatrix} \rho u_1 \\ \rho u_1^2 +      p \\ \rho u_1 u_2 \\ u_1(E + p) \end{Bmatrix}, 
\qquad \mathbf{F}_{2} = \begin{Bmatrix} \rho u_2 \\ \rho u_1 u_2  \\ \rho u_2^2 + p \\ u_2(E + p) \end{Bmatrix}.
$$
The viscous fluxes are given by
$$
\qquad \mathbf{F}_{v_1} = \begin{Bmatrix} 0 \\ \tau_{11} \\ \tau_{12}  \\ u_j \tau_{j1} + c_p \frac{\mu}{\text{Pr}} \frac{\partial T}{\partial x_1}  \end{Bmatrix}, 
\qquad \mathbf{F}_{v_2} = \begin{Bmatrix} 0 \\ \tau_{21} \\ \tau_{22}  \\ u_j \tau_{j2} + c_p \frac{\mu}{\text{Pr}} \frac{\partial T}{\partial x_2}  \end{Bmatrix}.
$$
We assume a Newtonian fluid, which leads to a viscous stress tensor of the form,
\begin{equation*}
\tau_{ij} = 2\mu S_{ij},
\end{equation*}
where,
\begin{equation*}
 S_{ij} = \frac{1}{2} \big( \frac{\partial u_i}{\partial \mathsf{x}_j} + \frac{\partial u_j}{\partial \mathsf{x}_i} \big) - \frac{1}{3} \frac{\partial      u_k}{\partial \mathsf{x}_i} \delta_{ij}.
\end{equation*}
The Navier-Stokes equations are closed with a constitutive relationship for a calorically perfect gas,
$$p = (\gamma - 1)( \rho E - \frac{1}{2} \rho u_1^2 - \frac{1}{2} \rho u_2^2 \big),$$
where $\gamma = 1.4$ is the heat-capacity ratio.

@@

\subsubsection{Description of FOM and generation of \spatialAcronym\ trial subspace}
The full-order model comprises a discontinuous-Galerkin discretization. The discretization 
is obtained by partitioning the domain into $100$ elements in the flow direction and $40$ elements 
in the wall-normal direction. The solution is represented to third order over each element using tensor product polynomials of order $p=2$, 
resulting in $36000$ unknowns for each conserved variable. Spatial discretization via the discontinuous Galerkin method yields a dynamical system 
of the form,
$$\frac{d \state}{dt} = \mass_{\text{DG}} \velocity_{\text{DG}} (\state),$$
where $\mass_{\text{DG}} \in \RR{N \times N}$ is the DG mass matrix and $\velocity_{\text{DG}}: \RR{N} \rightarrow \RR{N}$ is the DG velocity operator containing 
surface and volume integrals. By the definition of the FOM~\eqref{eq:FOM}, the dynamical system velocity is $\velocity = \mass_{\text{DG}} \velocity_{\text{DG}}$. 
Figure~\ref{fig:cav_mesh} shows the computational mesh. The 
DG method leverages the Rusanov flux at the cell interfaces~\cite{rusanov} and uses the first form of Bassi and Rebay~\cite{br1} for the viscous fluxes. Time integration 
is performed via a third-order strong stability preserving Runge-Kutta method~\cite{ssp_rk3} with a time-step size of $\Delta t = 0.001$.

@@

\subsubsection{Description of reduced-order models}
Hyper-reduced ROMs based on the Galerkin, collocated Least--Squares Petrov--Galerkin, and \methodAcronym\ approaches are 
again considered. Details on the implementation of the methods is as follows:
\begin{itemize}
\item \textit{Galerkin ROM with collocation}: The Galerkin ROM with collocation is obtained through Galerkin projection of the FOM. Unless 
otherwise noted, the  Galerkin ROM is evolved in time with the Crank--Nicolson time scheme at a time step of $\Delta t =0.1$.

\item \textit{LSPG ROM with collocation}: The LSPG ROM with collocation is built on top of the FOM discretization using the Crank--Nicolson scheme for temporal 
discretization. Unless otherwise noted, the time step for the LSPG ROM is $\Delta t  = 0.1$. As in the previous example, 
the Gauss--Newton method is used to solve the non-linear least-squares problem arising at each time instance. The Gauss--Newton 
algorithm is deemed converged when the gradient norm is less than $1e-4$. All Jacobians are again computed via automatic differentiation.

\item \textit{\methodAcronymROMs\ with collocation}: \methodAcronymROMs\ are solved via the direct method. The ROMs use the Crank--Nicolson time-discretization with a time-step size of 
$\Delta t = 0.1$. The Gauss--Newton algorithm is deemed converged when the gradient norm is less than $1e-4$. All Jacobians are computed via automatic differentiation. Uniform quadrature 
weights are used for evaluating the integral in~\eqref{eq:obj_gen_slab}.

@@

\subsubsection{Numerical results}
We first assess the performance of \methodAcronymROMs\ with varying window sizes. To this end, we consider \methodAcronymROMs\ with uniform window 
sizes of $\Delta T^n \equiv \Delta T = 0.2,0.5,1.0,$ and $2.0$, along with the Galerkin and LSPG ROMs. Results are first considered for basis \#1 as described in 
Table~\ref{tab:rom_basis_details}. For all ROMs, we evolve 
the solution for $t \in [0,100]$. This comprises the same time interval used to construct the trial subspace. First, Figure~\ref{fig:cav_results1a} depicts the evolution of the pressure at the bottom wall in the midpoint of the computational domain, while Figure~\ref{fig:cav_results1b} depicts the evolution of the normalized $\elltwo$ state error of the various reduced-order models. Both the hyper-reduced Galerkin and LSPG ROMs blow up/fail to converge within the first several time units. The \methodAcronymROM\ minimizing the residual over a window of size $\Delta T = 0.2$ also fails to converge. The \methodAcronymROMs\ that minimize the residual over window sizes of $\Delta T \ge 0.5$ are seen to all be stable and accurate; the pressure response is well characterized and the normalized state errors are again less than $10\%$. The most notable discrepancy between the \methodAcronymROM\ and FOM solutions is a slight phase difference.
